Stack Safety as a Security Property

### 2020-07-07 main.tex

The call stack is a perennially popular target for attacks
exploiting memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, many
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \apt{Mostly from
  nick;there could be more}\bcp{How about static techniques??  We decided
  these were also in-scope.}

But what {\em is} this ``stack safety'' of which we speak? Precise formal
definitions are lacking in the literature; instead, the notion of safety
is usually defined negatively, by enumerating the bad things that a safe system
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
We are only aware of one, quite recent, attempt to frame a positive definition.
\citet{Skorstengaard+19} define stack safety as
the conjunction of two properties, {\em local state encapsulation} and {\em
  well-bracketed control flow}.
They formalize these properties by defining an abstract capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this is an important step in the right direction, we believe
there is still room to craft a more abstract and less intensional
characterization of stack safety, so that we are able to answer
questions about other enforcement mechansims. That is the goal of this
paper.
\leo{I took a stab. I think it's still to dense, but captures the
  feeling of today's discussion I think}
\apt{still needs tuning}\bcp{Yes, I think we need to go into a
  little more detail on what their definition looks like in order to
  justify why an alternative definition might be valuable.  What
  questions does their presentation not answer? What uses does it not
  support? E.g., can it be enforced efficiently? (I guess so, by
  capabilities, though I think their capabilities are more powerful
  than what Cheri actually provides, while our policies are exactly
  what Nick needs.)  Why would it be good to look at stack safety
  through the lens of security?  (Because it is fundamentally a
  security mechanism?)}  \rb{BCP idea: contrast intrinsic
  vs. extrinsic property definition.  Remember to talk about spatial
  vs temporal safety}

Our main contribution in this paper is an alternative formal presentation of
local state encapsulation and well-bracketed control flow using the
conceptual and technical framework of language-based security.
%
The underlying notion behind local state encapsulation is the intention to
protect callers from callees: a callee should be not be able to access or
overwrite private data in the caller's stack frame. We formalize this
intuition as a security property with both confidentiality and integrity
components.
%
Intuitively, {\em stack confidentiality} promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  {\em Stack
  integrity} guarantees that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.  \bcp{We need to point out somewhere that we are only dealing with
stack safety, though we discuss (somewhere) how our property can be combined
with the heap safety property of [Azevedo et al]}\leo{Should this be
  here or in the assumptions as presented in the running example?}

The underlying notion behind well-bracketed control flow is
the need to protect the call structure itself: a callee should not be
allowed to jump to the middle of an arbitrary block of code either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). Well-bracketed control flow establishes safety
between functions: each instruction is uniquely owned by some
function, and all jumps between functions are either calls that enter
at a marked entry point or returns that re-enter just after their
associated call in the standard nested call structure.  \apt{clarify
  what we are contributing here.}

To phrase both aspects of stack safety as a security property requires
an explicit attacker model which, for this paper, is quite strong: a
callee can execute arbitrary code, and the enforcement mechanism
should prevent any stack safety violations. Moreover, even though
integrity and confidentiality are standard security properties,
applying them in the stack-safety setting raises a novel challenge:
the notion of ``attacker'' changes during execution. Not only does a
caller need to be protected from its callee, but the callee itself
needs to be protected from {\em its} callees. We address this
challenge by adjusting the integrity and confidentiality of different
stack frames at each call or return transition between
functions.
%\apt{connection to ``attacker'' still vague.}  \rb{Mention
%  to varying ``accessing power'' of the attacker?  The notion of
%  attacker changes on certain actions that are not externally
%  visible.} \bcp{Maybe the best way is to grab the bull by the horns:
%  ``If we're going to phrase stack safety as a security property, we
%  need to talk about an attacker model, and this is where the
%  technicalities get really interesting...''}

To provide evidence that our definitions capture the inituitive notion
of stack safety, we show that they rule out a range of exploitable bad
behaviors.\apt{``from the literature''??}  To assess whether they are
implementable, we compare them against the proposed capability-based
enforcement mechanism of \citet{Skorstengaard+19} \apt{we really
  should} and the existing tag-based enforcement mechanism of
\citet{DBLP:conf/sp/RoesslerD18}.\leo{Is this where we would rephrase this
  because we would only talk about it in related work?}

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for   this---''quasi-extensional''?}
  Abstract formal properties capturing the
  intuitive concepts of local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}) in a generic setting. These properties
  rule out common examples of stack memory abuse.
\item
  A case study of the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, based on hardware tags~(\cref{sec:enforcement}).
  This policy enforces our stack safety property, and also a (provably) stronger property that is easier to test.
  \apt{If we have no testing, do we even want to talk about the testing
    property?} \bcp{Not here; just in future work.  I guess we need to see
    whether we wind up with any results at all about testing.}
\item
  A refinement of the case study that considers the more efficient Lazy Tagging and Clearing policy
  of \citet{DBLP:conf/sp/RoesslerD18}~(\cref{sec:lazy}).  This policy does not enforce our property, but we give a
  variant of the policy that does enforce a (provably) weaker variant of our
  property.
%
  \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
    but the point we can make is still strong: To make the property easier
    to enforce, we can slightly weaken its guarantees; the weaker version is
    a bit more complicated to state but still retains the pragmatic
    protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
    depth-based tags, which do not enforce even our lazy policy. So we need to
    introduce a version of LT\&C that uses unique activation tags.}

  %\item As evidence that the property is not too strong\apt{But we add the lazy version because the strict version \emph{is} too strong, in some sense.}, we show that it is
%  realizible by
%  a previously proposed micropolicy enforcement
%  mechanism~(\Cref{sec:micropolicy}). \leo{And it says yes on a
%    substantial body of code}. We also present a variation of the
%  property + policy that is lazy, easier/more efficient to implement,
%  and provably weaker (\Cref{sec:lazy}).
%\item A provably stronger (inductive? step-by-step/stepwise) variant
%  of this property tuned for testing/verification.  We test the whole
%  thing with a cool framework. \bcp{Which section?}
\end{itemize}
\Cref{sec:relwork} discusses related work, and
\cref{sec:future} sketches directions for future work.
%
We begin with a running example that we refer
to throughout the paper (\cref{sec:running-example}) and some
technical preliminaries (\cref{sec:prelim}).

@@

A trace over a type $\TRACEELEMS$ is a nonempty, finite or infinite sequence of
$\TRACEELEMS$s. We will mostly be concerned with sequences of pairs of machine states
and observations, called machine traces and denoted $\machT \in \MOTRACE$,
and with sequences of triples of a machine state, a policy state, and an
observation, denoted $\MPT \in \MPOTRACE$ and referred to as MP-traces.

>>>>

The call stack is a perennial target for attacks
based on memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a menagerie of
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \apt{Mostly from
  nick;there could be more}\bcp{How about static techniques??  We decided
  these were also in-scope.}\bcp{... or are they?  I guess in any case we should
list any that we know about.}

But what {\em is} this ``stack safety'' of which we speak? Precise formal
definitions are lacking in the literature; instead, the notion
is usually defined negatively, by enumerating the bad things that a given
enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one---quite recent---attempt to frame a
positive definition:
\citet{Skorstengaard+19} define stack safety as
the conjunction of two properties, {\em local state encapsulation} and {\em
  well-bracketed control flow}.
They formalize these properties by defining an abstract capability-based machine
that ``satisfies [stack safety] by construction.''
%
Although this represents an important step in the right direction, we believe
there is still room to craft a more abstract, more generic characterization
of stack safety that is independent of capabilities or any other specific
enforcement
technique.  In other words, we seek formal criteria that can be applied
against \emph{any} stack-based system + enforcement mechanism to judge
whether or not the combination is actually stack-safe.

\iflater
\apt{another stab.}
\leo{I took a stab. I think it's still to dense, but captures the
  feeling of today's discussion I think}
\apt{still needs tuning}\bcp{Yes, I think we need to go into a
  little more detail on what their definition looks like in order to
  justify why an alternative definition might be valuable.  What
  questions does their presentation not answer? What uses does it not
  support? E.g., can it be enforced efficiently? (I guess so, by
  capabilities, though I think their capabilities are more powerful
  than what Cheri actually provides, while our policies are exactly
  what Nick needs.)  Why would it be good to look at stack safety
  through the lens of security?  (Because it is fundamentally a
  security mechanism?)}  \rb{BCP idea: contrast intrinsic
  vs. extrinsic property definition.  Remember to talk about spatial
  vs temporal safety}
\fi

The key contribution of this paper is a novel\bcp{generic?} presentation of
local state encapsulation and well-bracketed control flow using the
technical framework of language-based security.

The intuition behind \citet{Skorstengaard+19}'s\iflater\bcp{I forget how to
  remove the year and just have the author names here...}\fi use of the term {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should be not be able to access or overwrite private data
in the caller's stack frame. We formalize this intuition as a security
property with both confidentiality and integrity components.
%
We reflect this intuition in a formal
{\em stack confidentiality} property, which promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  Conversely, {\em stack
  integrity} promises that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.

\iflater
\bcp{We need to point out somewhere that we are only dealing with
stack safety, though we discuss (somewhere) how our property can be combined
with the heap safety property of [Azevedo et al]}\leo{Should this be
  here or in the assumptions as presented in the running
  example?}\bcp{Probably better there.}
\fi

The intuition behind \citet{Skorstengaard+19}'s {\em well-bracketed control
  flow} is 
the desire to protect the call graph itself: a callee should pass control
back to the caller only by returning in the standard way (transferring
control to the instruction following the call); it should 
not be
allowed to jump back into the caller's code at an arbitrary point, either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). Well-bracketed control flow establishes safety
between functions: each instruction is uniquely owned by some
function, and all jumps between functions are either calls that enter
at a marked entry point or returns that re-enter just after their
associated call in the standard nested call structure.  \apt{clarify
  what we are contributing here.}

To phrase these two aspects of stack safety in terms of security properties
requires an explicit attacker model. \bcp{This paragraph goes too fast: We
  need to start by asking, informally, who is being protected and who is the
attacker. Then we observe that real-world stack protection mechanisms tend
to imagine attack scenarios where the callee gets confused by something it
receives from the outside; to avoid getting into the details of how this
happens, we simply assume that the callee can execute arbitrary code.}
For simplicity, we choose one that is
fairly strong: when stating the protections that a function can rely on when
it makes a call, we assume that the callee can execute arbitrary
instructions until it returns.

\bcp{Again, here, a bigger transition, more setup, and a longer discussion
  are needed.  This is really the crux of what we are doing!}
Moreover, applying the standard language-based-security notions of
integrity and confidentiality in the stack-safety setting raises a novel
challenge:
the notion of ``attacker'' changes during execution. Not only does a
caller need to be protected from its callee, but the callee itself
needs to be protected from {\em its} callees. We address this
challenge by adjusting the integrity and confidentiality of different
stack frames at each call or return transition between
functions.  \bcp{that's way too terse!}
%\apt{connection to ``attacker'' still vague.}  \rb{Mention
%  to varying ``accessing power'' of the attacker?  The notion of
%  attacker changes on certain actions that are not externally
%  visible.} \bcp{Maybe the best way is to grab the bull by the horns:
%  ``If we're going to phrase stack safety as a security property, we
%  need to talk about an attacker model, and this is where the
%  technicalities get really interesting...''}

\bcp{This little paragraph is our entire explanation of the second and third
contributions!  Needs to be expanded considerably...}%
To provide evidence that our definitions capture the inituitive notion
of stack safety, we show that they rule out a range of exploitable bad
behaviors.\apt{``from the literature''??}  To assess whether they are
implementable, we compare them against
%proposed capability-based
%enforcement mechanism of \citet{Skorstengaard+19} \apt{we really
%  should} and
the existing tag-based enforcement mechanism of
\citet{DBLP:conf/sp/RoesslerD18}.
%\leo{Is this where we would rephrase this because we would only talk about it in related work?}

In summary, we offer the following contributions:\bcp{I would still like to
  see these points phrased in ``contribution normal form'': say exactly what
  we did for each one (it's a bit vague---''considers,'' etc.), and
  explicitly say why it is (a) useful, (b) challenging, and (c) novel.}
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for   this---''quasi-extensional''?}
  Abstract formal properties capturing the
  intuitive concepts of strict local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}) in a generic setting. These properties
  rule out common examples of stack memory abuse.
\item
  A case study of the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, based on hardware tags~(\cref{sec:enforcement}).
  This policy enforces our strict stack safety property, and also a (provably) stronger property that is easier to test.
  \apt{If we have no testing, do we even want to talk about the testing
    property?} \bcp{Not here; just in future work.  I guess we need to see
    whether we wind up with any results at all about testing.}
\item
  A refinement of the case study that considers the more efficient Lazy Tagging and Clearing policy
  of \citet{DBLP:conf/sp/RoesslerD18}~(\cref{sec:lazy}).  This policy does not enforce our strict
  local state encapsulation property,
  but a modified version of it using unique per-activation tags enforces
  a (provably) weaker lazy variant of the property.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

  %\item As evidence that the property is not too strong\apt{But we add the lazy version because the strict version \emph{is} too strong, in some sense.}, we show that it is
%  realizible by
%  a previously proposed micropolicy enforcement
%  mechanism~(\Cref{sec:micropolicy}). \leo{And it says yes on a
%    substantial body of code}. We also present a variation of the
%  property + policy that is lazy, easier/more efficient to implement,
%  and provably weaker (\Cref{sec:lazy}).
%\item A provably stronger (inductive? step-by-step/stepwise) variant
%  of this property tuned for testing/verification.  We test the whole
%  thing with a cool framework. \bcp{Which section?}
\end{itemize}
\Cref{sec:relwork} discusses related work;
\cref{sec:future} sketches directions for future work.
%
We begin with a running example that we refer
to throughout the paper (\cref{sec:running-example}) and some
technical preliminaries (\cref{sec:prelim}).

@@

A trace over a type $\TRACEELEMS$ is a nonempty, finite or infinite sequence of
$\TRACEELEMS$s. We will mostly be concerned with sequences of pairs of machine states
and observations, called machine traces and denoted $\machT \in \MOTRACE$,
and with sequences of triples of a machine state, a policy state, and an
observation, denoted $\MPT \in \MPOTRACE$ and referred to as MP-traces. 
observation, denoted $\MPT \in \MPOTRACE$ and referred to as MP-traces.
